---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This summarizes the steps taken to perform analysis in the data for the **Activity monitoring data**.  The variables cointained in this data set are: 

- **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
- **date**: The date on which the measurement was taken in YYYY-MM-DD format
- **interval**: Identifier for the 5-minute interval in which measurement was taken.

## Loading and preprocessing the data

We load the data that has been stored is in the local working directory. 
```{r loading}
activity<-read.csv("activity.csv")
activityN<-na.omit(activity)
activityN<-transform(activityN,interval=factor(interval))
head(activityN)
```


## What is mean total number of steps taken per day?
One can see the number of steps taken ecah dayin the following histogram:
```{r histMeanStepDay}
stepsperday<-with(activityN,tapply(steps,date,sum))
hist(stepsperday,main=" Histogram of number of steps ",xlab = "Total number of steps per day")
```

The mean and the median number of steps is computed for each day: 
```{r}
Average<-with(activityN,tapply(steps,date,mean,na.rm=TRUE))
Median1<-with(activityN,tapply(steps,date,median,na.rm=TRUE))
d0<-data.frame(date=names(Average),mean=Average);
d1<-data.frame(date=names(Average),median=Median1);
mrg<-merge(d0,d1,by="date")
mrg
```


## What is the average daily activity pattern?


We make a plot of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days.

```{r avNumStepDays}
d2<-with(activityN,tapply(steps,interval,mean))
d2<-data.frame(interval=as.integer(names(d2)),mean=d2);
plot(d2$interval,d2$mean, type="l",main="Average number of steps per interval",xlab="interval",ylab="average")
int<-which(d2$mean==max(d2$mean))
```

The maximum is achieved on the interval `r names(int)` in the position `r int` of the vector.

## Imputing missing values

```{r}
nm<-sum(is.na(activity))
nmd<-sum(!complete.cases(mrg))
```

The total number of missing values in the data set is `r nm`, but there are `r nmd` where no data point is given.  Therefore the strategy that will be used to fill in the missing values is to replace them by the mean for the respective interval.

A new data set is created with the values filled in.

```{r HistImputedStepsPerDay}
Iactivity=activity
for (i in 1:nrow(Iactivity)){
    if (is.na(Iactivity[i,]$steps))
    {
       inter<-Iactivity$interval[i]
       Iactivity[i,]$steps<-d2[d2$interval==inter,]$mean
    }
}
Istepsperday<-with(Iactivity,tapply(steps,date,sum))
hist(Istepsperday,main=" Histogram of number of steps (completed) ",xlab = "Total number of steps per day")

IAverage<-with(Iactivity,tapply(steps,date,mean))
IMedian1<-with(Iactivity,tapply(steps,date,median))
Id0<-data.frame(date=names(IAverage),mean.imputed=IAverage);
Id1<-data.frame(date=names(IAverage),median.imputed=IMedian1);
Imrg<-merge(Id0,Id1,by="date")
cmrg<-merge(mrg,Imrg,by="date")
cmrg
summary(cmrg)
```
 One can see that this has a tremendous impact on the number of steps, specially on the median values which is 0 for all other days where it not needed to ocmplete the data.


## Are there differences in activity patterns between weekdays and weekends?

We use the filled data frame but we create a new factor variable with the day of the week.
```{r}
Iactivity$day=weekdays(as.POSIXlt(Iactivity$date))
Iactivity$W<-rep("weekday",nrow(Iactivity))
Iactivity[Iactivity$day=="Sunday",]$W<-"weekend"
Iactivity[Iactivity$day=="Saturday",]$W<-"weekend"
Iactivity<-transform(Iactivity,W=factor(W))
gbwe<-subset(Iactivity,W=="weekend")
gbwd<-subset(Iactivity,W=="weekday")
```

We compute the average number of steps taken during the weekdays and the weekends and compare then in a plot.

```{r Compare}
gbwdd2<-with(gbwd,tapply(steps,interval,mean))
gbwdd2<-data.frame(interval=as.integer(names(gbwdd2)),mean=gbwdd2);
gbwed2<-with(gbwe,tapply(steps,interval,mean))
gbwed2<-data.frame(interval=as.integer(names(gbwed2)),mean=gbwed2);
par(mfrow=c(2,1))
plot(gbwed2$interval,gbwed2$mean, type="l",main="weekend",xlab=" ",ylab="avg num steps")
plot(gbwdd2$interval,gbwdd2$mean, type="l",main="weekday",xlab="interval",ylab="avg num steps")
```